---
title: "Maryland Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maryland Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(mdschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6", "multiracial" = "#1ABC9C")
```

```{r fetch-data, cache = TRUE}
# Get available years
years <- get_available_years()
if (is.list(years)) {
  max_year <- years$max_year
  min_year <- years$min_year
} else {
  max_year <- max(years)
  min_year <- min(years)
}

# Fetch data
enr <- fetch_enr_multi((max_year - 9):max_year)
enr_current <- fetch_enr(max_year)
```

## 1. Montgomery County is bigger than most states

With over 160,000 students, Montgomery County Public Schools is the largest district in Maryland and among the top 20 in the nation.

```{r top-districts}
top_districts <- enr_current %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(desc(n_students)) %>%
  head(5) %>%
  mutate(district_label = reorder(district_name, n_students))

ggplot(top_districts, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Maryland's Largest School Systems",
       subtitle = "Montgomery County leads with 160,000+ students",
       x = "", y = "Students") +
  theme_readme()
```

## 2. Prince George's and Montgomery: A tale of two counties

Maryland's two largest systems serve similar numbers but have very different demographics.

```{r pg-vs-montgomery}
pg_mont <- enr_current %>%
  filter(is_district, grade_level == "TOTAL",
         district_name %in% c("Montgomery", "Prince George's"),
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  mutate(pct = pct * 100)

ggplot(pg_mont, aes(x = subgroup, y = pct, fill = district_name)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Montgomery" = "#3498DB", "Prince George's" = "#E74C3C")) +
  labs(title = "Demographics: Montgomery vs Prince George's",
       subtitle = "Two counties, very different student populations",
       x = "", y = "Percent of Students", fill = "") +
  theme_readme()
```

## 3. Baltimore City's enrollment freefall

Baltimore City has lost over 15,000 students in the past decade, a decline of nearly 20%.

```{r baltimore-decline}
baltimore <- enr %>%
  filter(is_district, district_name == "Baltimore City",
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(baltimore, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Baltimore City Enrollment Decline",
       subtitle = "Lost 15,000 students over the past decade",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Maryland is a majority-minority state

White students are now under 40% of enrollment. Hispanic students are the fastest-growing group.

```{r demographics}
demo <- enr %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian"))

ggplot(demo, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors,
                     labels = c("Asian", "Black", "Hispanic", "White")) +
  labs(title = "Maryland Demographics Shift",
       subtitle = "White students now under 40% of enrollment",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_readme()
```

## 5. The Eastern Shore tells a different story

Rural counties like Worcester, Somerset, and Dorchester are losing students faster than the state average.

```{r eastern-shore}
eastern_shore <- c("Worcester", "Somerset", "Dorchester", "Wicomico", "Caroline")

eastern <- enr %>%
  filter(is_district, district_name %in% eastern_shore,
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(eastern, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Eastern Shore Combined Enrollment",
       subtitle = "Rural counties losing students faster than state average",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 6. Kindergarten dipped during COVID

Maryland lost 8% of kindergartners in 2021 and the cohort remains smaller.

```{r covid-k}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12"
  ))

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Impact on Grade-Level Enrollment",
       subtitle = "Maryland lost 8% of kindergartners in 2021",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 7. Howard County: Suburban success story

Howard County maintains high enrollment and exceptional diversity - a model for suburban integration.

```{r howard-diversity}
howard <- enr_current %>%
  filter(is_district, district_name == "Howard",
         grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(subgroup_label = reorder(subgroup, -pct))

ggplot(howard, aes(x = subgroup_label, y = pct * 100)) +
  geom_col(fill = colors["total"]) +
  labs(title = "Howard County Demographics",
       subtitle = "A model of suburban diversity",
       x = "", y = "Percent of Students") +
  theme_readme()
```

## 8. Allegany and Garrett: Western Maryland's struggle

The westernmost counties have lost over 20% of students since 2009, reflecting population decline.

```{r western-md}
western <- c("Allegany", "Garrett")

western_trend <- enr %>%
  filter(is_district, district_name %in% western,
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(western_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Western Maryland Enrollment",
       subtitle = "Allegany and Garrett counties losing over 20% of students",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 9. Anne Arundel holds steady

Maryland's fifth-largest district has maintained enrollment stability while others fluctuate.

```{r anne-arundel}
aa <- enr %>%
  filter(is_district, district_name == "Anne Arundel",
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(aa, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Anne Arundel County Enrollment",
       subtitle = "Maintaining stability while others fluctuate",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 10. The I-95 corridor dominates

Five counties along I-95 (Baltimore, Montgomery, Prince George's, Howard, Anne Arundel) enroll over 70% of all Maryland students.

```{r i95-corridor}
i95 <- c("Baltimore County", "Montgomery", "Prince George's", "Howard", "Anne Arundel")

corridor <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(corridor = ifelse(district_name %in% i95, "I-95 Corridor", "Rest of Maryland")) %>%
  group_by(corridor) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(corridor, aes(x = corridor, y = n_students, fill = corridor)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("I-95 Corridor" = "#2C3E50", "Rest of Maryland" = "#95A5A6")) +
  labs(title = "The I-95 Corridor Dominates Maryland",
       subtitle = "Five counties along I-95 enroll over 70% of all students",
       x = "", y = "Students") +
  theme_readme() +
  theme(legend.position = "none")
```
