---
title: "Maryland Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maryland Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(mdschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "highlight" = "#E67E22", "secondary" = "#95A5A6",
            "decline" = "#E74C3C", "growth" = "#27AE60", "blue" = "#3498DB")
```

```{r fetch-data, cache = TRUE}
enr <- fetch_enr_multi(2016:2024, use_cache = TRUE)
enr_current <- fetch_enr(2024, use_cache = TRUE)

# Helper function to get unique district totals
get_district_totals <- function(df) {
  df %>%
    filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
    select(end_year, district_name, n_students) %>%
    distinct()
}

# Helper function to get unique state totals
get_state_totals <- function(df) {
  df %>%
    filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
    select(end_year, n_students) %>%
    distinct()
}
```

## 1. Montgomery County is Maryland's largest district with 155,000 students {#montgomery-county}

Montgomery County Public Schools enrolls more students than entire states like Wyoming or Vermont. It is one of the top 20 largest school districts in the nation.

```{r story-01-top-districts}
top_districts <- get_district_totals(enr_current) %>%
  arrange(desc(n_students)) %>%
  head(5) %>%
  mutate(district_label = reorder(district_name, n_students))

stopifnot(nrow(top_districts) > 0)

top_districts %>%
  select(district_name, n_students)

ggplot(top_districts, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Maryland's Largest School Systems, 2024",
       subtitle = "Montgomery County leads with 155,000 students",
       x = "", y = "Students") +
  theme_readme()
```

## 2. Maryland enrollment has been remarkably stable since 2016 {#statewide-enrollment}

Maryland has maintained roughly 855,000-877,000 students over the past 9 years. A brief dip during COVID in 2020-2021 recovered by 2022, and enrollment has held steady near 859,000.

```{r story-02-state-trend}
state_trend <- get_state_totals(enr) %>%
  arrange(end_year)

stopifnot(nrow(state_trend) > 0)

state_trend %>%
  select(end_year, n_students)

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020, y = max(state_trend$n_students) + 2000,
           label = "COVID", color = "red", size = 3) +
  scale_y_continuous(labels = comma) +
  labs(title = "Maryland Statewide Enrollment, 2016-2024",
       subtitle = "Remarkably stable at ~859,000 students",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 3. Baltimore City lost nearly 5,000 students since 2016 {#baltimore-city-decline}

Baltimore City enrollment has declined 6.3% from 77,866 in 2016 to 72,995 in 2024. The decline has been steady, with a slight uptick in 2024. Population loss and suburban migration are driving forces.

```{r story-03-baltimore-decline}
baltimore <- get_district_totals(enr) %>%
  filter(district_name == "Baltimore City") %>%
  arrange(end_year)

stopifnot(nrow(baltimore) > 0)

baltimore %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(baltimore, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["decline"]) +
  geom_point(size = 3, color = colors["decline"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Baltimore City Enrollment Decline",
       subtitle = "Lost nearly 5,000 students (-6.3%) from 2016 to 2024",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Kindergarten dropped 10% in 2020 and never fully recovered {#covid-kindergarten}

COVID hit kindergarten hardest. Maryland lost 10.3% of kindergartners in the 2019-20 school year as families delayed enrollment. By 2024, kindergarten enrollment is still 8.5% below its 2019 peak, suggesting some students shifted permanently out of public schools.

```{r story-04-covid-k}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "12")) %>%
  select(end_year, grade_level, n_students) %>%
  distinct() %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12"
  ))

stopifnot(nrow(k_trend) > 0)

k_trend %>%
  filter(grade_level == "K") %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020, y = max(k_trend$n_students, na.rm = TRUE) + 1000,
           label = "COVID", color = "red", size = 3) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Impact on Grade-Level Enrollment",
       subtitle = "Kindergarten dropped 10.3% in 2020 and has not recovered",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 5. The Eastern Shore has barely changed in a decade {#eastern-shore}

Five Eastern Shore counties (Worcester, Somerset, Dorchester, Wicomico, and Caroline) collectively serve about 33,000 students. Unlike the dramatic declines seen elsewhere, the Eastern Shore has been remarkably stable since 2016.

```{r story-05-eastern-shore}
eastern_shore <- c("Worcester", "Somerset", "Dorchester", "Wicomico", "Caroline")

eastern <- get_district_totals(enr) %>%
  filter(district_name %in% eastern_shore) %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") %>%
  arrange(end_year)

stopifnot(nrow(eastern) > 0)

eastern %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(eastern, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Eastern Shore Combined Enrollment",
       subtitle = "Five rural counties hold steady at ~33,000 students",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 6. Western Maryland is shrinking: Garrett lost 12% since 2016 {#western-maryland}

Allegany and Garrett counties in Appalachian Maryland have lost 7% and 12% of students respectively since 2016. These rural mountain communities face similar population challenges to Appalachian communities nationwide.

```{r story-06-western-md}
western <- c("Allegany", "Garrett")

western_trend <- get_district_totals(enr) %>%
  filter(district_name %in% western) %>%
  arrange(district_name, end_year)

stopifnot(nrow(western_trend) > 0)

western_trend %>%
  group_by(district_name) %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1)) %>%
  filter(!is.na(change)) %>%
  select(district_name, end_year, n_students, change, pct_change)

ggplot(western_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Western Maryland Enrollment Decline",
       subtitle = "Garrett lost 12.2%, Allegany lost 7.0% since 2016",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 7. Anne Arundel grew 4% while nearby Baltimore County shrank {#anne-arundel}

Anne Arundel County has quietly added 3,200 students since 2016, growing 4.1%. Meanwhile, neighboring Baltimore County lost over 2,300 students. The Annapolis-area county benefits from military families at Fort Meade and its proximity to both DC and Baltimore.

```{r story-07-anne-arundel}
aa <- get_district_totals(enr) %>%
  filter(district_name == "Anne Arundel") %>%
  arrange(end_year)

stopifnot(nrow(aa) > 0)

aa %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(aa, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["growth"]) +
  geom_point(size = 3, color = colors["growth"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Anne Arundel County Enrollment Growth",
       subtitle = "Added 3,200 students (+4.1%) since 2016",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 8. The I-95 corridor enrolls 61% of all Maryland students {#i95-corridor}

Five counties along I-95 (Baltimore County, Montgomery, Prince George's, Howard, and Anne Arundel) enroll 526,000 of Maryland's 859,000 students. This 61% concentration reflects the state's suburban population center between DC and Baltimore.

```{r story-08-i95-corridor}
i95 <- c("Baltimore County", "Montgomery", "Prince George's", "Howard", "Anne Arundel")

corridor <- get_district_totals(enr_current) %>%
  mutate(corridor = ifelse(district_name %in% i95, "I-95 Corridor", "Rest of Maryland")) %>%
  group_by(corridor) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(corridor) == 2)

corridor %>%
  mutate(pct = round(n_students / sum(n_students) * 100, 1))

ggplot(corridor, aes(x = corridor, y = n_students, fill = corridor)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("I-95 Corridor" = "#2C3E50", "Rest of Maryland" = "#95A5A6")) +
  labs(title = "The I-95 Corridor Dominates Maryland Education",
       subtitle = "Five counties along I-95 enroll 61% of all students",
       x = "", y = "Students") +
  theme_readme() +
  theme(legend.position = "none")
```

## 9. Frederick County grew 16% -- fastest in the state {#frederick-growth}

Frederick County has added over 6,300 students since 2016, a 15.8% increase. Located between the DC suburbs and western Maryland, Frederick attracts families seeking more affordable housing while maintaining access to the DC job market.

```{r story-09-frederick-growth}
frederick <- get_district_totals(enr) %>%
  filter(district_name == "Frederick") %>%
  arrange(end_year)

stopifnot(nrow(frederick) > 0)

frederick %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(frederick, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["growth"]) +
  geom_point(size = 3, color = colors["growth"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Frederick County Enrollment Growth",
       subtitle = "DC exurbs driving 15.8% growth since 2016",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 10. Baltimore County vs Baltimore City: Divergent paths {#baltimore-comparison}

While Baltimore City lost nearly 5,000 students (-6.3%), Baltimore County also shrank but from a higher base. The county surrounds but is entirely separate from the city. Both are losing students, though at different rates.

```{r story-10-baltimore-comparison}
baltimore_both <- get_district_totals(enr) %>%
  filter(district_name %in% c("Baltimore City", "Baltimore County")) %>%
  arrange(district_name, end_year)

stopifnot(nrow(baltimore_both) > 0)

baltimore_both %>%
  group_by(district_name) %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1)) %>%
  filter(!is.na(change)) %>%
  select(district_name, end_year, n_students, change, pct_change)

ggplot(baltimore_both, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Baltimore City" = "#E74C3C", "Baltimore County" = "#3498DB")) +
  labs(title = "Baltimore City vs Baltimore County",
       subtitle = "Both declining, from different starting points",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 11. Howard County plateaued after years of growth {#howard-plateau}

Howard County grew from 54,348 in 2016 to 57,508 in 2019, then flattened. The county is now one of the few large districts in Maryland where enrollment has stopped growing, possibly reflecting the limits of available housing.

```{r story-11-howard-plateau}
howard <- get_district_totals(enr) %>%
  filter(district_name == "Howard") %>%
  arrange(end_year)

stopifnot(nrow(howard) > 0)

howard %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(howard, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["blue"]) +
  geom_point(size = 3, color = colors["blue"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Howard County Enrollment Plateau",
       subtitle = "Growth stalled after 2019 at ~56,000 students",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 12. Charles County: Southern Maryland's steady growth {#charles-county}

Charles County is the largest district in Southern Maryland and has added 1,483 students (+5.8%) since 2016. The county serves as a bedroom community for DC-area workers seeking affordable housing south of the capital.

```{r story-12-charles-county}
charles <- get_district_totals(enr) %>%
  filter(district_name == "Charles") %>%
  arrange(end_year)

stopifnot(nrow(charles) > 0)

charles %>%
  filter(end_year %in% c(min(end_year), max(end_year))) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(charles, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Charles County Enrollment",
       subtitle = "Southern Maryland's largest system: steady +5.8% growth",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 13. Grade 9 is 30% larger than grade 12 {#grade-structure}

Maryland has 77,465 ninth-graders but only 63,844 twelfth-graders -- a 21% drop. This pattern reflects both students leaving the system before graduation and the state's grade promotion policies.

```{r story-13-grade-structure}
grade_data <- enr_current %>%
  filter(is_state, subgroup == "total_enrollment", grade_level != "TOTAL") %>%
  select(grade_level, n_students) %>%
  mutate(grade_num = case_when(
    grade_level == "K" ~ 0,
    TRUE ~ as.numeric(grade_level)
  )) %>%
  arrange(grade_num)

stopifnot(nrow(grade_data) > 0)

grade_data %>%
  select(grade_level, n_students) %>%
  arrange(desc(n_students))

ggplot(grade_data, aes(x = factor(grade_level, levels = c("K", sprintf("%02d", 1:12))),
                       y = n_students)) +
  geom_col(fill = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Maryland Enrollment by Grade Level, 2024",
       subtitle = "Grade 9 (77K) is 30% larger than Kindergarten (60K)",
       x = "Grade", y = "Students") +
  theme_readme()
```

## 14. Montgomery County peaked in 2019 and has been declining {#montgomery-decline}

Maryland's largest district reached 160,587 students in 2019, then lost nearly 6,000 students by 2024. This 3.6% decline in the state's flagship district signals broader suburban enrollment pressure.

```{r story-14-montgomery-trend}
montgomery <- get_district_totals(enr) %>%
  filter(district_name == "Montgomery") %>%
  arrange(end_year)

stopifnot(nrow(montgomery) > 0)

montgomery %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

ggplot(montgomery, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["decline"]) +
  geom_point(size = 3, color = colors["decline"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Montgomery County: From Growth to Decline",
       subtitle = "Lost nearly 6,000 students since 2019 peak",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 15. Small counties face existential challenges {#small-counties}

Kent County has just 1,591 students, Somerset has 2,619, and Garrett has 3,193. These are among the smallest school districts in the mid-Atlantic region, making it difficult to offer diverse programs and maintain facilities.

```{r story-15-small-counties}
small_counties <- c("Kent", "Somerset", "Garrett")

small_trend <- get_district_totals(enr) %>%
  filter(district_name %in% small_counties) %>%
  arrange(district_name, end_year)

stopifnot(nrow(small_trend) > 0)

small_trend %>%
  filter(end_year == max(end_year)) %>%
  select(district_name, n_students) %>%
  arrange(n_students)

ggplot(small_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Maryland's Smallest School Systems",
       subtitle = "Kent, Somerset, and Garrett face enrollment pressure",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

```{r session-info}
sessionInfo()
```
